---
title: "Optimal selection of priority areas for biological surveillance"
author: "Carlos Zambrana-Torrelio"
date: "1/3/2017"
output: html_document
---

```{r Setup, download data from IUCN and Joint Research Center, include=FALSE}
library(letsR)
library(tidyverse)
library(stringr)
library(raster)
library(cshapes)
library(rprojroot)

P <- rprojroot::find_rstudio_root_file
# Load functions to be used
source(P('R/gvp_function.R'))
source(P('R/make_grid.R'))
knitr::opts_chunk$set(echo = TRUE)
dir.create(P('data/zip'))



# Download terrestrial mammals data from AWS. Source: IUCN redlist
if(!file.exists(P("data/zip/IUCN_MAMMALS_04_19_17.zip"))) {
  download.file("https://s3.amazonaws.com/gvp-shapefiles/IUCN_MAMMALS_04_19_17/IUCN_MAMMALS_04_19_17.zip", destfile = P("data/zip/IUCN_MAMMALS_04_19_17.zip"))
}
if(file.exists(P("data/zip/IUCN_MAMMALS_04_19_17.zip")) &
   !dir.exists(P("data/mammals_data"))) {
  unzip(P("data/zip/IUCN_MAMMALS_04_19_17.zip"),
        exdir=P("data/mammals_data/"))
}

# Download accesibility data from AWS. Source: http://forobs.jrc.ec.europa.eu/products/gam/
# Check if .zip file exists in directory if not, downloads from AWS
if(!file.exists(P("data/zip/access/access_50k.zip"))) {
  download.file("https://s3.amazonaws.com/ehaspatial/access_50k.zip", destfile = P("data/zip/access_50k.zip"))
}

# If accessibility zip file exist, but the data is not uncompressed then unzip
if(file.exists(P("data/zip/access_50k.zip")) &
   !dir.exists(P("data/access"))) {
  unzip(P("data/zip/access_50k.zip"),
        exdir=P("data/access/"))
}
```

## Here we create RDS files that are saved and can be used later in the marxan_files_light without having to read shapefiles again

```{r Read mammals shapefile and create a presence-absence matrix at different degrees: 1, 2, 5, 10}

# Read shapefile
terr = shapefile(P("/data/mammals_data/IUCN_MAMMALS_04_19_17/TERRESTRIAL_MAMMALS.shp"), verbose = T)
terr@data$binomial = str_replace(terr@data$binomial, " ", "_") #replace space in binomial name with an underscore
#terr = subset(terr, PRESENCE == 1 & LEGEND != 'Extinct' | LEGEND != "Possibly Extinct" | LEGEND != 'Not Mapped')


=======
terr = subset(terr, binomial != 'Ursus_maritimus') #remove polar bears (much of their extent is marine)

#Select only name, presence, and origin variables 
sub_terr = terr[, c(2:4)]
sub_terr$binomial <- as.factor(sub_terr$binomial)

```
### If we want to get presence/absence per Order rather than species we need to subset the mammal sp dataframe and then apply lets.presab
```{r merge polygons}

order = unique(terr$order_name)

terr_order_2deg <- lapply(order, function(x){
  t <- terr[terr$order_name == x,] #subset to order
  t <- t[,c(2:4)] #choose only binomial, presence, origin columns
  t$binomial = as.factor(t$binomial) 
  t <- lets.presab(t, resol = 2, presence = 1, origin = 1)
  r <- t[[2]] #get the Richness raster
  r[r > 0] = 1 #consider all counts as presences 
  names(r) = x # assing layer name to be the same as order name
  return(r)
})

# We need to convert this list of rasters into a raster stack & reproject

terr_order_2deg <- stack(unlist(terr_order_2deg)) 
terr_order_2dg_eck <- projectRaster(terr_order_2deg, crs = eck_proj, method = 'ngb') #reprojection

saveRDS(terr_order_2dg_eck, P('data/output/terr_order_2dg_eck.RDS'))

```


```{r reprojection}
# Reproject Eckert IV
eck_proj = CRS('+proj=eck4 +lon_0=0 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs')

#Create a presence-absence matrix of species' geographic ranges whithin a grid and save file as RDS
# 10 degrees
terr_pam_all_10dg = lets.presab(sub_terr, resol = 10, presence = 1, origin = 1)
saveRDS(terr_pam_all_10dg, P('data/output/terr_pam_all_10dg.RDS'))
# 5 degrees
terr_pam_all_5dg = lets.presab(sub_terr, resol = 5, presence = 1, origin = 1)
saveRDS(terr_pam_all_5dg, P('data/output/terr_pam_all_5dg.RDS'))
# 2 degrees
terr_pam_all_2dg = lets.presab(sub_terr, resol = 2, presence = 1, origin = 1)
saveRDS(terr_pam_all_2dg, P('data/output/terr_pam_all_2dg.RDS'))
# 1 degree
terr_pam_all_1dg = lets.presab(sub_terr, resol = 1, presence = 1, origin = 1)
saveRDS(terr_pam_all_1dg, P('data/output/terr_pam_all_1dg.RDS'))

terr_pam_all_0.2dg = lets.presab(sub_terr, resol = 0.2, presence = 1, origin = 1)
saveRDS(terr_pam_all_0.2dg, P('data/output/terr_pam_all_0.2dg.RDS'))


# Projection of prescence and absence of order_name
terr_pam_all_2dg_order = lets.presab(sub_terr_order, resol = 2, presence = 1, origin = 1)
saveRDS(terr_pam_all_2dg_order, P('data/output/terr_pam_all_2dg_order.RDS'))
# 1 degree
terr_pam_all_1dg_order = lets.presab(sub_terr_order, resol = 1, presence = 1, origin = 1)
saveRDS(terr_pam_all_1dg_order, P('data/output/terr_pam_all_1dg_order.RDS'))


### by order ###
terr_pam_rodentia_2dg = lets.presab(sub_rodentia, resol = 2, presence = 1, origin = 1)
saveRDS(terr_pam_rodentia_2dg, P('data/output/terr_pam_rodentia_2dg.RDS'))
terr_pam_rodentia_1dg = lets.presab(sub_rodentia, resol = 1, presence = 1, origin = 1)
saveRDS(terr_pam_rodentia_1dg, P('data/output/terr_pam_rodentia_1dg.RDS'))

terr_pam_2dg_chiroptera = lets.presab(sub_chiroptera, resol = 2, presence = 1, origin = 1)
saveRDS(terr_pam_chiroptera_2dg, P('data/output/terr_pam_chiroptera_2dg.RDS'))


crop_BGD = crop(terr_pam_all_0.2dg[[2]], BGD)
crop_BGD2 <- projectRaster(crop_BGD, crs = projection(pu_pol_BGD))
plot(crop_BGD2)
plot(pu_pol_BGD, add = T)


```

```{r RasterStack with prescence/absence mammals}
# Get shapefiles for each individual species
#Rasterize matrix element of the lets.presab output
#Get matrix from terr_pam_all_2dg
terr_pam_all_mat_0.2dg = terr_pam_all_0.2dg[[1]]
# set xy coordinates
xy_0.2dg = terr_pam_all_mat_0.2dg[,1:2]
# species presence absence matrix 
presabs_0.2dg = terr_pam_all_mat_0.2dg[,-c(1:2)]
#Rasterize each column of matrix using terr_pam_all_1dg raster as template
terr_pam_all_ras_1dg = terr_pam_all_1dg[[2]]
ras_0.2dg = apply(terr_pam_all_mat_0.2dg, 2, function(x) rasterize(xy_0.2dg, terr_pam_all_ras_1dg, field = x))
#create raster stack
terr_pam_all_0.2dg_ras = stack(unlist(ras_0.2dg))
saveRDS(terr_pam_all_0.2dg_ras, P('data/output/terr_pam_all_0.2dg_ras.RDS'))



# Get shapefiles for each order: Rodentia
#Rasterize matrix element of the lets.presab output
#Get matrix from terr_pam_all_2dg
terr_pam_rodentia_mat_2dg = terr_pam_2dg_rodentia[[1]]
# set xy coordinates
xy_2dg_rodent = terr_pam_rodentia_mat_2dg[,1:2]
# species presence absence matrix 
presabs_2dg_rodent = terr_pam_rodentia_mat_2dg[,-c(1:2)]
#Rasterize each column of matrix using terr_pam_all_1dg raster as template
terr_pam_rodentia_ras_1dg = terr_pam_rodentia_1dg[[2]]
ras_2dg_rodent = apply(terr_pam_rodentia_mat_2dg, 2, function(x) rasterize(xy_2dg_rodent, terr_pam_rodentia_ras_1dg, field = x))
#create raster stack
terr_pam_rodentia_2dg_ras = stack(unlist(ras_2dg_rodent))
saveRDS(terr_pam_rodentia_2dg_ras, P('data/output/terr_pam_rodentia_2dg_ras.RDS'))

```


## Generate access_2dg (polygon with access costs in minutes)

```{r cost, echo=FALSE}
# Read access_50k raster
access = raster(P('/data/access_50k/acc_50k'))
access_2dg = raster_to_low_res_grid(access, terr_pam_all_2dg[[2]])
saveRDS(access_2dg, P('data/output/access_2dg.RDS')) 

# Read country map
world = cshp(date=as.Date("2016-01-01"))

```

## We change the projection of the raster files because we want to create Planning Units (PUs) with an hexagonal form and to do that we need to define a projection in meters

```{r Reprojection of files}
# Define Coordinate Reference System arguments (CRS) (units will now be in meters)
eck_proj = CRS('+proj=eck4 +lon_0=0 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs')
# Project values of old rasters into new raster with new CRS, (ngb=nearest neighbor)
terr_pam_all_10dg_eck = projectRaster(terr_pam_all_10dg[[2]], crs = (eck_proj), method = 'ngb')
terr_pam_all_5dg_eck = projectRaster(terr_pam_all_5dg[[2]], crs = (eck_proj), method = 'ngb')
terr_pam_all_2dg_eck = projectRaster(terr_pam_all_2dg[[2]], crs = (eck_proj), method = 'ngb')
terr_pam_all_1dg_eck = projectRaster(terr_pam_all_1dg[[2]], crs = (eck_proj), method = 'ngb')
```

## We load the world map so we can generate a grid of planning units in hexagons (pu_pol_hex) and a data frame of the grid (pu_pol_hex_Df)

```{r Load world polygon and create hexagon grid}
# Read and load a polygon with the world borders
data(wrld_simpl)
# Exclude antartica from polygon
wrld_simpl = subset(wrld_simpl, NAME != 'Antarctica')
# Eckert IV projection - define units in meters
world_eckert <- CRS('+proj=eck4 +lon_0=0 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs') %>% 
  spTransform(wrld_simpl, .)

### COSTA RICA ###
CRI = getData(name  = "GADM", country = "CRI", level = 1)
e = drawExtent()
CRI = crop(CRI, e)
CRI_eckert <- CRS('+proj=eck4 +lon_0=0 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs') %>% 
  spTransform(CRI, .)
# Generate grid of polygons of 100km2 each
pu_pol_CRI <- make_grid(CRI_eckert, type="hexagonal", cell_area = 100000000, clip = TRUE)
#####

### BANGLADESH - PU 100 Km2 ###
BGD = getData(name  = "GADM", country = "BGD", level = 0)
plot(BGD)
e = drawExtent()
BGD = crop(BGD, e)
BGD_eckert <- CRS('+proj=eck4 +lon_0=0 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs') %>% 
  spTransform(BGD, .)
# Generate grid of polygons of 100km2 each
pu_pol_BGD <- make_grid(BGD_eckert, type="hexagonal", cell_area = 500000000, clip = TRUE)
#####


# Generate grid of hexagons using make_grid function with each hexagon = 50,000 km2
pu_pol_hex <- make_grid(world_eckert, type="hexagonal" ,cell_area = 50000000000, clip = TRUE)
saveRDS(pu_pol_hex, P('data/output/pu_pol_hex.RDS'))
# Generate an index of the number of hexagons in pu_pol_hex
id_hex = tibble(id_hex = c(1:5113)) # be careful with this index because it does not update to the number of hexagons generated by pu_pol_hex
# Transform pu_pol_hex into a data frame
pu_pol_hex_df <-  SpatialPolygonsDataFrame(pu_pol_hex, data=id_hex)
saveRDS(pu_pol_hex_df, P('data/output/pu_pol_hex_df.RDS'))


# Generate grid of hexagons using make_grid function with each hexagon = 100,000 km2
pu_pol_hex_100k <- make_grid(world_eckert, type="hexagonal" ,cell_area = 100000000000, clip = TRUE)
saveRDS(pu_pol_hex_100k, P('data/output/pu_pol_hex_100k.RDS'))
# Generate an index of the number of hexagons in pu_pol_hex_100k
id_hex_100k = tibble(id_hex_100k = c(1:3102)) # be careful with this index because it does not update to the number of hexagons generated by pu_pol_hex
# Transform pu_pol_hex into a data frame
pu_pol_hex_100k_df <-  SpatialPolygonsDataFrame(pu_pol_hex_100k, data=id_hex_100k)
saveRDS(pu_pol_hex_100k_df, P('data/output/pu_pol_hex_100k_df.RDS'))

# Generate grid of hexagons using make_grid function with each hexagon = 200,000 km2
pu_pol_hex_200k <- make_grid(world_eckert, type="hexagonal" ,cell_area = 200000000000, clip = TRUE)
saveRDS(pu_pol_hex_200k, P('data/output/pu_pol_hex_200k.RDS'))
# Generate an index of the number of hexagons in pu_pol_hex_100k
id_hex_200k = tibble(id_hex_200k = c(1:1968)) # be careful with this index because it does not update to the number of hexagons generated by pu_pol_hex
# Transform pu_pol_hex into a data frame
pu_pol_hex_200k_df <-  SpatialPolygonsDataFrame(pu_pol_hex_200k, data=id_hex_200k)
saveRDS(pu_pol_hex_200k_df, P('data/output/pu_pol_hex_200k_df.RDS'))
```


Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
